% !TEX encoding = UTF-8 Unicode
\acknowledgments

First, I would like to thank my advisor, Carole Ober. Thanks for providing a rigorous but nurturing environment to grow into a scientist. Throughout my PhD, I have grown to respect and look up to you even more which I didn't think was possible. Thanks for providing the perfect amount of mentorship for us to succeed. Thanks to Dan Nicolae for teaching me so much and encouraging me to think outside of the box and meet with him to discuss all sorts of progress and ideas. Thanks to my committee for taking the time out of their busy schedules to provide me with lots of direction and support, both in one on one and in committee meetings: Yoav Gilad, John Novembre, and Vincent Lynch.  

My decision to pursue Genetics in college came about after my program advisor, Sarah Dempsey, rightly so, refused to let me take both genetics and cell biology lab courses in the same semester. After crying in her office, I was forced to make a decision and chose the one I enjoyed learning about more and told myself I'd worry about the pre-optometry requirements later. It was through these genetics courses that I fell in love with the field and met Daniel Rokhsar. Dan took a chance on me and let me work in his lab for a senior honors thesis during my last year at UC Berkeley. He has since encouraged me and been a supporter of me and my science. Many thanks to Eric Edsinger and Jess Lyons for teaching me how to science, and Adam Session, Therese Mitros, Simon Prochnik, and Jessen Bredeson for teaching me how to code and work with genomic data. Special thanks to my undergraduate friends in the lab who are now pursuing their own graduate degrees: Cindy Ha and Konnor La. 

With only one year of science research under my belt, I was admitted into one graduate university - University of Chicago. Thank you to the GGSB faculty who fought for me to come here. I've had the great opportunity to work with some strong amazing women in science, Nancy Cox, Elizabeth McNally, and Hae Kyung Im, among many others, who have helped me along my graduate career from the beginning, from writing me strong letters of recommendation, supporting conference attendance, enrolling me in R Master classes with Hadley Wickham, and convincing me to become active in the science community on Twitter - all of which have contributed to make me the scientist I am today.

I would like to thank the opportunities and outside activities that have helped keep me sane and motivated. Thanks to my fellow volunteers at the Museum of Science and Industry and Matt, Rene, and Rudy for giving me the opportunity to teach adults and children about DNA for the past three years. I had the privilege of answering random questions, having engaging discussions, and correcting misconceptions which I find important as a scientist and more crucial in our current political climate.

It should be acknowledged that none of this happened without any struggles. It's hard to forget the teaching assistant during my first year of graduate school, who instead of helping me, told me I didn't know anything and said the faculty questioned my presence in graduate school; or the peers who forget about the female computational biologists - when we can all learn from each other.

I would not have made it through graduate school without the support of my friends and colleagues. Katie Igartua and Michelle Stein, thanks for welcoming me to the lab, teaching me a lot of what I know today and leaving the bar quite high. Thanks to the past, current, and future members of the lab who have helped make it the great environment that it is and provided help with everything all these years. Thanks to my cohort of Molecular Bioscience students who have gone through the same things with me since Day 1- especially Lindsey Montefiori, Ana Beringer, Lindsey Mao, Erin Boyle Anderson, Chris Katanski, Phil McGilvray, Chris Craddock, and Beth Pollard. Thanks to the older GGSB students who welcomed me in when I was the only GGSB student in my cohort : Kenneth Barr, Nick VanKuren, John Blischak, Bryce van de Geijn, Aashish Jha, Ziyue Gao, Talia Karasov, Patrick Landback, and Jason Pitt. Thanks for all the advice through the years, all sorts of computational help, and most importantly the invitation to play soccer with The Repressors. The GGSB students who have made our program exceptional and especially those I am lucky to call my friends: UnJin Lee, Charlie Lang, Leila Reyes Ruiz, Briana Mittleman, and Linsin Smith. Thanks to the Human Genetics students who included me in their department, especially Erin Fry, Ittai Eres, Katie Mika, Lauren Blake, and Alex Gileta. 
Thanks to my many friends from other programs and departments who I have come to be really close with: Dana Gilmore \& Brendan Horton, Amelia Joslin \& Andrew Tremain, Keston \& Ayelet Aquino Michaels, Larischa de Wet. Also Kyle Cron, Sangman Kim, Toufic Mayassi, Chris Stamper, Jaime Chao, Chanie Howard, Ryan Duncombe, and Steven Erickson, Colles Price, Jason Torres, Kevin Lei, Joel Smith, and Ali Ekrem Yesilkanal. 

I have to thank all the administrators and advisors that made my graduate career path go smoothly as well as help me achieve my dream job, especially Sue Levison, Briana Konnick, and Abby Stayart. 

Thanks to the fuzzy butts that have provided emotional support and furry comfort along the way: Efe, Audrey, Penny (dog), Meatball, Darth Vader, Penny (cat), and Panda.

Thanks to Alan Chang for being my partner these past four years and making every day brighter. Thanks for exploring Chicago in food and drink with me and being my favorite adventure buddy. I can't imagine having gone through any of this without your encouragement and love by my side.

Thanks to my entire family for their loving support. Thanks to my Aunt Peymaneh and Uncle Mahmood for sending me letters full of love, Halloween packages full of my favorite candy, and a whole slough of Christmas gifts for when I wasn't home to celebrate. Thanks to my grandma for always asking to FaceTime with me - but only for a short time to let me get back to work- and for sending me too many raw pistachios, pomegranates from her backyard, and homemade Persian sweets I miss during Persian New Year every year for the past five years. Thanks to my sisters, my best friends, who will talk to me early in the morning (California is two hours behind Chicago) and let me rant about any and everything. None of this would have been possible without the support of my parents, who didn't question my decision to move halfway across the country for the past five years but continued to visit (albeit only in the summer) and tell me how proud they are of me. Thanks for pushing me as hard as you did and taking care of me all of these years. 
